---
title: "From Scholarship Records to One Row per Student"
subtitle: "STAT 392 · A worked data-wrangling example"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set() sets defaults for all code chunks:
# echo = TRUE displays the code; message = FALSE hides startup messages.
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
# options() changes how R displays results. Here, tibble.width = Inf
# asks printed tibbles to show all columns.
options(tibble.width = Inf)
```

## What are we trying to produce?

We begin with several scholarship-process records for each student. We want
**one row per student**, with a column listing their associated scholarships
and a column summarizing their request status.

The data below are entirely fictional and included in this file. They use
the relevant column names and kinds of values from the project data, but
contain no actual student records. This example covers **one application
year**. If you combine years, group by both student and application year.

Two columns contain different kinds of information:

| Column | How we use it here |
|:--|:--|
| `Request Status` | One value per record; used in the main example. |
| `Submission Status` | Several semicolon-separated values; explored in the optional section. |

This walkthrough assumes a specific definition of **“furthest,”** shown in
Step 5. Depending on your definition, you may use a different ranking or
follow slightly different data-wrangling steps.

Open this file in RStudio. Run the chunks **in order**, inspecting each
intermediate result, or click **Knit** to generate an HTML walkthrough.

## 0. Load the packages

- `library(tidyverse)` loads the installed tidyverse packages so their
  functions are available in this R session.

```{r load-packages}
# Load the tidyverse, which includes the packages used in this example.
library(tidyverse)
```

The pipe, `%>%`, passes a result to the next operation. Names containing
spaces, such as `` `Process Name` ``, need backticks when used in R code.

## 1. Create a small dummy dataset

Each row is a process record associated with a student. A student has a
general application record and additional named-scholarship records.
The example has ten records for three students. Each student–process
combination occurs once here.

What the functions do:

- `tibble()` builds a table from the named columns supplied inside it.
- `c()` combines several values into one vector—a sequence of values for a
  column. The first value in each column belongs to the first row, the second
  to the second row, and so on.

`NA` marks a missing value. Writing the object name `scholarship_data` on its
own displays the table.

```{r create-dummy-data}
# All names, scholarship titles, and award amounts below are invented.
scholarship_data <- tibble(
  fake_first = c(
    "Avery", "Avery", "Avery",
    "Jordan", "Jordan", "Jordan",
    "Maya", "Maya", "Maya", "Maya"
  ),
  fake_last = c(
    "Chen", "Chen", "Chen",
    "Lopez", "Lopez", "Lopez",
    "Patel", "Patel", "Patel", "Patel"
  ),
  `Process Name` = c(
    "2025 Community Scholarship Application",
    "River Valley Scholarship 2025",
    "Community Leadership Award 2025",
    "2025 Community Scholarship Application",
    "River Valley Scholarship 2025",
    "STEM Futures Scholarship 2025",
    "2025 Community Scholarship Application",
    "Community Leadership Award 2025",
    "STEM Futures Scholarship 2025",
    "First Generation Scholars Award 2025"
  ),
  `Submission Status` = c(
    "Submitted; Assigned",
    "Submitted; Submitted; Complete",
    "Submitted; Complete; Complete; Assigned; Complete",
    "Submitted; Assigned",
    "Submitted; Submitted; Complete",
    "Submitted; Submitted; Complete",
    "Submitted; Assigned; Assigned",
    "Submitted; Submitted; Complete",
    "Submitted; Complete; Complete; Assigned; Assigned; Complete",
    "Submitted; Submitted; Complete"
  ),
  `Request Status` = c(
    "Application Submitted", "Denied", "Follow Up(s) Assigned",
    "Application Submitted", "Denied", "Denied",
    "Application Submitted", "Denied", "Follow Up(s) Assigned", "Denied"
  ),
  `Amount Awarded` = c(
    NA, NA, 2500,
    NA, NA, NA,
    NA, NA, 4000, NA
  )
)

# Inspect the starting records before changing anything.
scholarship_data
```

Notice that `Submission Status` can contain several values, while
`Request Status` contains one value. We will not split `Submission Status`
until the optional section.

## 2. Identify students consistently

For these fictional records, first name plus last name identifies a student.
This is appropriate only when the identifier is known to be consistent and
unique; names are not generally reliable identifiers in other datasets.

What the functions do:

- `mutate()` adds the new `student_id` column while keeping the existing
  columns and rows.
- `str_c(fake_first, fake_last, sep = " ")` joins the first and last name
  **within each row**, with a space between them.
- `distinct()` keeps the specified columns and one copy of each unique
  combination. Here, it reduces the repeated student names to a roster.

```{r identify-students}
# Combine the two name fields to identify the same student across rows.
scholarship_data <- scholarship_data %>%
  mutate(student_id = str_c(fake_first, fake_last, sep = " "))

# Save a roster before filtering any records, so no student disappears
# simply because they have only a general application record.
student_roster <- scholarship_data %>%
  distinct(student_id, fake_first, fake_last)

# Inspect the three students represented in the data.
student_roster
```

## 3. Separate general applications from named scholarships

The general application is not itself a named scholarship. We exclude it
from the scholarship list, but retain it when summarizing request statuses.
For real data, first inspect the process names and identify the appropriate
general-application label or labels.

What the functions do:

- `mutate()` adds `is_general_application`, a column of `TRUE`/`FALSE` values.
- `filter()` keeps rows for which its condition is `TRUE`. Here,
  `!is_general_application` keeps rows that are not general applications.
- `select()` keeps the columns we want to display so we can inspect the
  classification easily.

The comparison `==` checks whether two values are equal. The operator `!`
means “not,” reversing `TRUE` and `FALSE`.

```{r identify-scholarship-records}
# Mark the general application using its exact fictional process name.
scholarship_data <- scholarship_data %>%
  mutate(
    is_general_application =
      `Process Name` == "2025 Community Scholarship Application"
  )

# Keep named-scholarship records in a separate object.
scholarship_records <- scholarship_data %>%
  filter(!is_general_application)

# Check the classification before relying on it.
scholarship_data %>%
  select(student_id, `Process Name`, is_general_application)
```

## 4. Combine scholarship names into one column

What the functions do:

- `group_by(student_id)` tells subsequent operations to work separately
  within each student; it does not yet reduce the number of rows.
- `summarise()` produces one row per student from the summary we define.
- `unique()` removes repeated scholarship names within that student's group.
- `sort()` arranges those names alphabetically.
- `str_c(..., collapse = "; ")` combines **all the names in the group** into
  one text value, separated by semicolons. Unlike `sep` in Step 2, `collapse`
  joins values across the vector rather than joining columns within a row.

The argument `.groups = "drop"` returns an ungrouped table, so later
operations do not automatically continue working student by student.

```{r combine-scholarships}
# Combine the distinct named scholarships associated with each student.
scholarship_list <- scholarship_records %>%
  group_by(student_id) %>%
  summarise(
    scholarships_considered = str_c(
      sort(unique(`Process Name`)),
      collapse = "; "
    ),
    .groups = "drop" # Return an ordinary, ungrouped table.
  )

# Each row should now represent one student.
scholarship_list
```

We use the label `scholarships_considered` here. Confirm what a process record
represents before describing this list as scholarships a student explicitly
applied for.

## 5. Define a request-status priority rule

In this example, “furthest” means the status with the highest rank in this
table. These ranks tell R which label to keep when a student has several
different request statuses.

| Request status | Rank |
|:--|--:|
| Application Submitted | 1 |
| Denied | 2 |
| Follow Up(s) Assigned | 3 |

What the functions do:

- `mutate()` adds `request_status_rank` to each original record.
- `case_when()` checks the conditions in order and returns the value after
  `~` for the first condition that matches. Here, it translates status text
  into a number; `TRUE ~ NA_real_` leaves other labels with a missing numeric
  rank.
- `select()` displays the student, process, original status, and new rank
  together so we can check the translation.

```{r rank-request-status}
# Assign a numeric rank to each recognized status.
status_data <- scholarship_data %>%
  mutate(
    request_status_rank = case_when(
      `Request Status` == "Application Submitted" ~ 1,
      `Request Status` == "Denied" ~ 2,
      `Request Status` == "Follow Up(s) Assigned" ~ 3,
      TRUE ~ NA_real_ # Leave missing or unrecognized statuses unranked.
    )
  )

# Check that the ranks match the rule above.
status_data %>%
  select(student_id, `Process Name`, `Request Status`, request_status_rank)
```

The following check lists labels that did not receive a rank. An empty
table is expected for this dummy dataset.

- `is.na()` checks whether a rank is missing, returning `TRUE` or `FALSE`.
- `filter()` keeps only rows with missing ranks.
- `distinct()` shows each unranked status label once.

```{r inspect-unranked-statuses}
# Identify labels that the priority rule does not cover.
status_data %>%
  filter(is.na(request_status_rank)) %>%
  distinct(`Request Status`)
```

## 6. Select one request status per student

What the functions do:

- `is.na()` identifies missing ranks; `!` reverses the result, identifying
  ranks that are present.
- `filter()` keeps rows with a recognized, nonmissing rank.
- `group_by(student_id)` organizes the next step separately for each student.
- `slice_max()` keeps the row with the largest rank in each group.
  `n = 1` requests one row, and `with_ties = FALSE` keeps just one if several
  rows share the maximum.
- `ungroup()` removes the grouping after the selection is finished.
- `select()` keeps the ID, status, and rank. The notation `new_name = old_name`
  also renames the selected columns.

Tied maximum ranks have the same status label here, so selecting one is
sufficient for this status summary.

```{r summarize-request-status}
# Select the highest-priority recognized status across each student's records.
furthest_request_status <- status_data %>%
  filter(!is.na(request_status_rank)) %>%
  group_by(student_id) %>%
  slice_max(
    order_by = request_status_rank,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  select(
    student_id,
    furthest_request_status = `Request Status`,
    furthest_request_status_rank = request_status_rank
  )

# Inspect the one-status-per-student result.
furthest_request_status
```

A student with no recognized status will be absent from this intermediate
table. Joining to the full roster later preserves that student and gives
them a missing summary status. If only some statuses are unrecognized,
the summary reflects only the recognized ones; resolve those labels first.

## 7. Keep outcome information separately

An overall status hides differences across scholarship records. Separate
flags preserve whether any denial or follow-up was recorded and whether any
positive award amount was recorded.

In this example, each nonmissing amount represents a separate award that
appears once. **Check that assumption before summing real records.** Repeated
records or a student-level total repeated across scholarships would lead
to double counting.

What the functions and conditional do:

- `group_by(student_id)` makes each summary use one student's records.
- `summarise()` creates one row per student, with the outcome columns defined
  inside it.
- `any()` asks whether **at least one** value meets a condition—for example,
  whether at least one award amount is positive. Here, `na.rm = TRUE` ignores
  missing values while checking.
- `is.na()` identifies the missing award amounts.
- `all()` asks whether **every** value is `TRUE`. Thus,
  `all(is.na(...))` checks whether all of a student's award amounts are missing.
- `if (...) { ... } else { ... }` chooses between two results: return `NA`
  when all amounts are missing; otherwise calculate the sum.
- `sum()` adds the recorded amounts. Here, `na.rm = TRUE` omits missing
  amounts from the calculation.

Again, `.groups = "drop"` returns an ungrouped summary table.

```{r summarize-outcomes}
# Summarize the observed evidence across each student's records.
student_outcomes <- scholarship_data %>%
  group_by(student_id) %>%
  summarise(
    had_denial = any(`Request Status` == "Denied", na.rm = TRUE),
    had_follow_up = any(
      `Request Status` == "Follow Up(s) Assigned",
      na.rm = TRUE
    ),
    has_recorded_award = any(`Amount Awarded` > 0, na.rm = TRUE),

    # Preserve NA if every amount is missing; do not convert missing to zero.
    total_recorded_award = if (all(is.na(`Amount Awarded`))) {
      NA_real_
    } else {
      sum(`Amount Awarded`, na.rm = TRUE)
    },
    .groups = "drop"
  )

# Inspect the outcome flags and recorded amounts.
student_outcomes
```

Here, `FALSE` means the feature was **not observed in these records**.
For example, `has_recorded_award = FALSE` does not establish that a student
received no award when the amounts are missing. A sum of observed amounts
is also not necessarily a complete total if some amounts are missing.

## 8. Join the student-level summaries

Each summary now has one row per student. Starting from the roster retains
students who lack named-scholarship rows.

What the functions do:

- `left_join()` adds columns from another table by matching `student_id`.
  It keeps every row of the table on the left and fills unmatched columns
  with `NA`. Here, each summary has at most one match per student, so the
  joins preserve one row per student.
- `select()` chooses the main result columns to display. The full
  `student_summary` object still contains all the joined columns.

```{r join-summaries}
# Add each summary to the complete student roster.
student_summary <- student_roster %>%
  left_join(scholarship_list, by = "student_id") %>%
  left_join(furthest_request_status, by = "student_id") %>%
  left_join(student_outcomes, by = "student_id")

# Display the main result: one row per student.
student_summary %>%
  select(
    student_id,
    scholarships_considered,
    furthest_request_status,
    has_recorded_award,
    total_recorded_award
  )
```

Check the result against the original ten rows:

| Student | Named scholarships | Selected request status | Recorded award total |
|:--|--:|:--|--:|
| Avery Chen | 2 | Follow Up(s) Assigned | 2500 |
| Jordan Lopez | 2 | Denied | NA |
| Maya Patel | 3 | Follow Up(s) Assigned | 4000 |

Jordan's missing total remains `NA`. Avery and Maya each have a denial in
another scholarship record, even though their selected status is
Follow Up(s) Assigned. You can see this in `student_outcomes`.

## Optional: summarize the values inside Submission Status

If “furthest” refers to the values inside `Submission Status`, first split
the semicolon-separated field before ranking and summarizing.

### 9. Give each original record an ID

We need a record ID to remember which row each status came from after
splitting the field.

What the functions do:

- `mutate()` adds a `record_id` column without removing any existing columns.
- `row_number()` numbers the rows in their current order: 1, 2, 3, and so on.
  We assign these numbers before splitting, so each expanded row will retain
  its original record ID.

```{r identify-original-records}
# Number the original records before expanding them.
submission_long <- scholarship_data %>%
  mutate(record_id = row_number())
```

### 10. Split the field at each semicolon

For example, `Submitted; Submitted; Complete` becomes three rows with
the same `record_id`.

What the functions do:

- `separate_rows()` splits the status field at each semicolon (`sep = ";"`)
  and puts each piece on its own row, repeating the other column values.
- `mutate()` replaces the status column with its cleaned version.
- `str_trim()` removes spaces at the beginning and end of each status value.
- `filter(record_id <= 3)` keeps the expanded rows belonging to the first
  three original records for inspection.
- `select()` displays only the record ID, student, process, and status.

```{r split-submission-status}
# Expand each semicolon-separated field into individual status values.
submission_long <- submission_long %>%
  separate_rows(`Submission Status`, sep = ";") %>%
  mutate(`Submission Status` = str_trim(`Submission Status`))

# Inspect the first three original records after expansion.
submission_long %>%
  filter(record_id <= 3) %>%
  select(record_id, student_id, `Process Name`, `Submission Status`)
```

All other columns are repeated when a row is expanded. **Do not sum award
amounts in this expanded table:** an award would be counted once per status
value instead of once per original record.

### 11. Define an ordering

For this optional example, assume the ordering Submitted = 1, Complete = 2,
Assigned = 3.

What the functions do:

- `mutate()` adds a numeric rank to each expanded status row.
- `case_when()` matches each status label to the corresponding number.
  Its final `TRUE ~ NA_real_` assigns a missing rank to any other value.

```{r rank-submission-status}
# Apply the chosen ordering to each individual status value.
submission_long <- submission_long %>%
  mutate(
    submission_status_rank = case_when(
      `Submission Status` == "Submitted" ~ 1,
      `Submission Status` == "Complete" ~ 2,
      `Submission Status` == "Assigned" ~ 3,
      TRUE ~ NA_real_
    )
  )
```

### 12. Select one status per original record

First summarize within `record_id`, not within student. This preserves a
separate result for each original process record.

What the functions do:

- `is.na()` identifies missing ranks; `!` identifies the nonmissing ones.
- `filter()` keeps only rows with a recognized status rank.
- `group_by(record_id)` collects the expanded rows from each original record
  into their own group.
- `slice_max()` selects the largest rank within each record.
  `n = 1, with_ties = FALSE` keeps a single row even if a status is repeated.
- `ungroup()` removes the record-level grouping after the selection.
- `select()` keeps the identifiers, status, and rank, and gives the last two
  columns descriptive names using `new_name = old_name`.

```{r summarize-submission-by-record}
# Select the highest-ranked recognized value within each original record.
furthest_submission_by_record <- submission_long %>%
  filter(!is.na(submission_status_rank)) %>%
  group_by(record_id) %>%
  slice_max(
    order_by = submission_status_rank,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  select(
    record_id,
    student_id,
    `Process Name`,
    furthest_submission_status = `Submission Status`,
    furthest_submission_status_rank = submission_status_rank
  )

# Inspect the simplified status for each original record.
furthest_submission_by_record
```

Records with no recognized values are absent from this intermediate table.
To retain all original records, join this result back to a roster of record
IDs. If a student–scholarship combination has multiple original records,
this is still a record-level table, not yet one row per student–scholarship.

### 13. Optionally summarize again by student

Only take this additional step if the research question needs one summary
for the student across all process records. This example includes general
application records; filter those out first if the intended summary should
describe named scholarships only.

What the functions do:

- `group_by(student_id)` groups the record-level results by student.
- `slice_max()` selects the largest record-level rank for each student;
  `n = 1, with_ties = FALSE` keeps one row.
- `ungroup()` removes the student grouping once the selection is complete.
- `select()` keeps only the student ID and the selected status and rank.
- `left_join()` matches this result to the full roster by `student_id`,
  retaining students who have no recognized submission status.

```{r summarize-submission-by-student}
# Select the highest-ranked record-level status within each student.
furthest_submission_by_student <- furthest_submission_by_record %>%
  group_by(student_id) %>%
  slice_max(
    order_by = furthest_submission_status_rank,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  select(
    student_id,
    furthest_submission_status,
    furthest_submission_status_rank
  )

# Keep all students, even if none of their submission values were recognized.
student_roster %>%
  left_join(furthest_submission_by_student, by = "student_id")
```

All three students receive Assigned under this rule because each has
Assigned in the general application record.

## Adapting this example

Before using the pattern on the project data, decide:

- **Unit of analysis:** one student, one student per year, or one
  student–scholarship combination? Group by the corresponding identifiers.
- **Meaning of a process:** does it indicate a submitted application,
  eligibility, consideration, or something else? Label the list accordingly.
- **Meaning of repeated records and amounts:** do records represent distinct
  awards, repeated exports, or multiple stages? Resolve this before adding
  amounts or removing duplicates.

Keep the original records and the intermediate summaries. A student-level
summary is useful, but it no longer shows which status belongs to which
scholarship. Return to the detailed records when that relationship matters.
